---
title: 'EDS 223: assignment 2'
author: "Colleen McCamy"
date: "2022-10-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## introduction

The following exercises are modified from [Chapters 3](https://geocompr.robinlovelace.net/attr.html), [4](https://geocompr.robinlovelace.net/spatial-operations.html), [5](https://geocompr.robinlovelace.net/geometry-operations.html) of Geocomputation with R by Rovin Lovelace. Each question lists the total number of points. The breakdown of points can be found at the end of each instruction in parentheses. A general grading rubric can be found on the [course website](https://ryoliver.github.io/EDS_223_spatial_analysis/assignments.html).

Please update "author" to list your first and last names and any collaborators (e.g. Ruth Oliver, Friend1, Friend2).

**Due by midnight on Saturday 2022-10-22**

## prerequisites

```{r load, include=TRUE, message=FALSE, warning=FALSE}
# add required packages here

library(sf)
library(spData)
library(tmap)
library(tidyverse)
library(rmapshaper)
library(rmapshaper)
library(smoothr)
```

## question 1

##### 5 points

Find the states that: (2.5)\
- belong to the West region\
- have an area below 250,000 square kilometers \#### Units are needed!!\
- and greater than 5,000,000 residents in 2015

```{r include=TRUE}

#loading in the data and mutating AREA to numeric
us_data <- spData::us_states
class(us_data$AREA)
us_data <- us_data |> 
  mutate(AREA = as.numeric(AREA))

# filtering the us_data for the western regions
us_western <- us_data |> 
  filter(REGION == "West")
#filtering the us_data for the area below 250,000
us_area250000_below <- us_data |> 
  filter(AREA < 250000)
# filtering the states by population 
us_pop_greater_5mil <- us_data |> 
  filter(total_pop_15 > 5000000)

#combining the three filtering pieces
all_three_attributes <- us_data |> 
  filter(REGION == "West") |> 
  filter(AREA < 250000) |> 
  filter(total_pop_15 > 5000000)

# ANSWER:
print(paste(all_three_attributes$NAME, "is in the Western region of the US, has an area below 250,000 km^2, and had a population greater than 5 million in 2015."))

```

What was the total population of the US in 2015? (2.5)

```{r include=TRUE}

#finding the sum
total_pop_2015 <- sum(us_data$total_pop_15)

#ANSWER:
print(paste0("The total population of the contiguous US in 2015 was ",total_pop_2015, " people."))

```

## question 2

##### 5 points

Create a new variable named "us_states_stats" by adding variables from "us_states_df" to "us_states". (3)

-   Which function did you use and why? (0.5)

-   Which variable is the key variable in both datasets? (0.5)

-   What is the class of the new object? (0.5)

-   Why does us_states_df have 2 more rows than us_states? (0.5)

```{r include=TRUE}

class(us_states)
class(us_states_df)

# edited the column name for states
us_states_df <- us_states_df |> 
  mutate(NAME = state)
#merged the data
us_states_stats <- merge(us_states, us_states_df, by = "NAME")

#ANSWER:
print(paste("I used the merge() function as the class for us_states is an sf object and data.frame and the us_states_df class is a data.frame. I noticed that using merge excluded Alaska and Hawaii from the new object as they didn't have spatial geometries associated with them. If I wanted to keep the data for Alaska and Hawaii, I could have used a left join."))

#ANSWER: 
print(paste("The state name is the key variable between both datasets."))

class(us_states_stats)
#ANSWER:
print(paste("The class of the new object us_states_stats is an sf object and data.frame."))

#ANSWER: 
print(paste("The us_states_df included data for 'Alaska' and 'Hawaii' while the the 'us_states'only included in the contiguous United States. Thus, in merging the data with the code below, the data for Alaska and Hawaii in the 'us_states_df' was omitted."))

```

## question 3

##### 10 points

Make a map of the percent change in population density between 2010 and 2015 in each state. Map should include a legend that is easily readable. (7)

```{r include=TRUE}
#adding population density difference in the dataframe to map
us_states_map <- us_states_stats |> 
  mutate(AREA = as.numeric(AREA)) |> 
  mutate(pop_den_15 = total_pop_15/AREA) |> 
  mutate(pop_den_10 = total_pop_10/AREA) |> 
  mutate(pop_den_diff = pop_den_15 - pop_den_10)

#mapping the results
tm_shape(us_states_map) +
  tm_fill(col = "pop_den_diff", 
          title = "Population Difference",
          palette = "Spectral") +
  tm_layout(legend.outside = TRUE, 
            main.title = "US States by Difference in Total Population Density from 2010 to 2015",
            main.title.size = 1) +
  tm_borders(alpha = 0.7)
```

In how many states did population density decrease? (3)

```{r include=TRUE}

#filtering for population difference less than 0
pop_decrease <- us_states_map |> 
  filter(pop_den_diff < 0)
#seeing how many rows are in the dataframe which will tell us how many states saw a decrease in population
nrow(pop_decrease)

#ANSWER:
print(paste(nrow(pop_decrease), "states, Michigan and Rhode Island, had a decrease in population from 2010 to 2015."))
```

## question 4

##### 10 points

How many of New Zealand's high points are in the Canterbury region? (5)

```{r include=TRUE}

#storing the data in the environment
nz <- nz
nz_height <- nz_height

# creating an object for Canterbury data
canterbury <- nz |>
  filter(Name == "Canterbury")

#figuring out all of the points inside canterbury similar to select
canterbury_heights <- nz_height[canterbury,]

#determining the number of points  
nrow(canterbury_heights)

#ANSWER:
print(paste(nrow(canterbury_heights), "of New Zealand's highest points are in the Canterbury region."))
```

Which region has the second highest number of "nz_height" points? And how many does it have? (5)

```{r include=TRUE}

#combining the data for nz_height and nz ojects
region_height <- st_join(nz_height, nz)

# ordering the data based on regions which have heights
region_height_order <- region_height |> 
  count(Name, sort = TRUE)
 
#ANSWER:
print(paste("The second highest number of 'nz_height' points is the", region_height_order$Name[2], "with", region_height_order$n[2], "points within the nz_height data."))
```

## question 5

##### 15 points

Create a new object representing all of the states the geographically intersect with Colorado.(5)\
Hint: use the "united_states" dataset. The most concise way to do this is with the subsetting method "[".\
Make a map of the resulting states. (2.5)

```{r include=TRUE}

us_states <- us_states

# creating object of Colorado
co <- us_states |>
  filter(NAME == "Colorado")

# creating a TRUE/FALSE list of states that intersect with Colorado
co_intersect <- st_intersects(x = us_states, y = co)
co_logical <- lengths(co_intersect) > 0

# filtering the list based on the logical TRUE/FALSE statement
co_intersect_true <- us_states[co_logical, ]

# mapping the data
tm_shape(co_intersect_true) +
  tm_polygons(col = "#d7f7f7", 
              border.col = "#056e6e") +
  tm_layout(main.title = "US States Intersecting with Colorado",
            main.title.color = "#056e6e",
            main.title.position = "center") +
  tm_text("NAME", 
          col = "#056e6e",
          size = 1.5)
```

Create another object representing all the objects that touch (have a shared boundary with) Colorado and plot the result.(5)\
Hint: remember you can use the argument op = st_intersects and other spatial relations during spatial subsetting operations in base R).\
Make a map of the resulting states. (2.5)

```{r include=TRUE}

# creating a TRUE/FALSE list of states that touch Colorado
co_touch <- st_touches(x = us_states, y = co)
co_logical_touch <- lengths(co_touch) > 0

# filtering the list based on the logical TRUE/FALSE statement
co_touch_true <- us_states[co_logical_touch, ]

# mapping the data
tm_shape(co_touch_true) +
  tm_polygons(col = "#d9bdd7", 
              border.col = "#453444") +
  tm_layout(main.title = "US States Touching Colorado",
            main.title.color = "#453444",
            main.title.position = "center") +
  tm_text("NAME", 
          col = "#453444",
          size = 1.5)

```

## question 6

##### 10 points

Generate simplified versions of the "nz" dataset. Experiment with different values of keep (ranging from 0.5 to 0.00005) for **ms_simplify()** and dTolerance (from 100 to 100,000) **st_simplify()**. (5)

Map the results to show how the simplification changes as you change values.(5)

```{r include=TRUE}

# ----- using ms_simplify() 
nz_ms_0.5 <- ms_simplify(nz, 0.5)
nz_ms_0.0005 <- ms_simplify(nz, 0.0005)
nz_ms_0.005 <- ms_simplify(nz, 0.005)
nz_ms_0.05 <- ms_simplify(nz, 0.05)

#creating maps 
ms_0 <-tm_shape(nz) +
  tm_polygons() +
  tm_layout(main.title = "No Simplification")
ms_1 <- tm_shape(nz_ms_0.5) +
  tm_polygons() +
    tm_layout(main.title = "0.5 Simplification")
ms_2 <- tm_shape(nz_ms_0.005) +
  tm_polygons() +
  tm_layout(main.title = "0.005 Simplication")
ms_3 <- tm_shape(nz_ms_0.0005) +
  tm_polygons() +
  tm_layout(main.title = "0.0005 Simplification")
ms_4 <- tm_shape(nz_ms_0.05) +
  tm_polygons() +
  tm_layout(main.title = "0.05 Simplication")

# joining the maps into one figure
ms_map_comparison <- tmap_arrange(ms_0, ms_1, ms_4, ms_2, ms_3)
ms_map_comparison

# ----- using st_simplify()
nz_st_100 <- st_simplify(nz, preserveTopology = FALSE, dTolerance = 100)
nz_st_1000 <- st_simplify(nz, preserveTopology = FALSE, dTolerance = 1000)
nz_st_10000 <- st_simplify(nz, preserveTopology = FALSE, dTolerance = 10000)
nz_st_36000<- st_simplify(nz, preserveTopology = FALSE, dTolerance = 36000)

#creating the maps with the simplifications above
st_0 <- tm_shape(nz) +
  tm_polygons() +
  tm_layout(main.title = "No Simpliciation")
st_1 <- tm_shape(nz_st_100) +
  tm_polygons() +
    tm_layout(main.title = "100 Simplification")
st_2 <- tm_shape(nz_st_1000) +
  tm_polygons() +
    tm_layout(main.title = "1000 Simplification")
st_3 <- tm_shape(nz_st_10000) +
  tm_polygons() +
    tm_layout(main.title = "10000 Simplification")
st_4 <- tm_shape(nz_st_36000) +
  tm_polygons() +
    tm_layout(main.title = "36000 Simplification")

#combining all the maps in one figure
st_map_comparison_2 <- tmap_arrange(st_0, st_1, st_2, st_3, st_4)
st_map_comparison_2
```

## question 7

##### 10 points

How many points from the "nz_height" dataset are within 100km of the Canterbury region?

```{r include=TRUE}

# creating a buffer zone of Canterbury 
canterbury_buffer <- st_buffer(canterbury, 100000)

# fiitering the nz_height using the 
buffer_zone <- nz_height[canterbury_buffer, ]
nrow(buffer_zone)

#ANSWER: 
print(paste("There are", nrow(buffer_zone), "points from the nz_height dataset that are within 100km of the Canterbury region."))

```

## question 8

##### 15 points

Find the geographic centroid of the country of New Zealand. How far is it from the geographic centroid of Canterbury?

```{r include=TRUE}

#combining the geometries
nz_geom <- st_combine(nz) 
nz_geom_all <- st_union(nz)

#finding the centroid of NZ
nz_centroid <- st_centroid(nz_geom_all)

# finding the centroid of canterbury
canterbury_centroid <- st_centroid(canterbury)

#mapping it to see the centroid of NZ (blue) and canterbury (red)
tm_shape(nz_geom) +
  tm_polygons() +
  tm_shape(nz_centroid) +
  tm_dots("blue") +
  tm_shape(canterbury_centroid) +
  tm_dots("red")

#calculating the distance between the two centroids
st_distance(nz_centroid, canterbury_centroid)

#ANSWER: 
print(paste("The distnance between the geographic centroid of New Zeland and Canterbury is", ceiling(st_distance(nz_centroid, canterbury_centroid)), "km^2"))
```
